---
title: "Extra examples"
author:
  - R. Fisher
  - S.K. Wilson
  - S.M. Sin
  - A.C. Lee
  - Dr Tim J. Langlois

output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extra examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(FSSgam)
library(mgcv)
library(MuMIn)
library(tidyverse)
require(car)
require(doBy)
require(gplots)
require(RColorBrewer)
library(tibble)
library(dplyr)
library(tidyr)
```

## Overview

He we show some simpler examples of using FSSgam for different response distributions and particularly show the use with binomial data, which none of the original case studies were able to do.

## R packages and data

The dataset used here is for a large range of coral species identified as points scores on individual images, with a small number of potential predictors.

```{r packages}
library(FSSgam)
library(mgcv)
library(MuMIn)
require(car)
require(doBy)
require(gplots)
require(RColorBrewer)
library(tibble)
library(dplyr)
library(tidyr)

dat <- coral_data
str(dat)
```

## Example showing use of uGamm to allow fitting with gamm4 

We set up the necessaru predictor vectors

```{r preds}
cat.preds=c("Survey","bleach.pres","dredge.pres","dhw.fact")
null.vars=c("Site")
cont.preds=c("av.wave","Depth")
```

Now do some basic data manipulation to get rid of NA's and unused columns

```{r wrangle}
use.dat=na.omit(dat[,c(null.vars,cat.preds,cont.preds,"allcoral","totalpoints")])
use.dat$successes=use.dat$allcoral
use.dat$failures=use.dat$totalpoints-use.dat$allcoral
use.dat$trials=use.dat$totalpoints
```

Now we fit the test.fit model for all coral, with total points as trials, and run the FSSgam approach.

```{r fit, include=FALSE}
Model1=uGamm(cbind(successes,failures)~s(Depth,k=4,bs='cr'),
              family=binomial(), random=~(1|Site),
             data=use.dat,
             lme4=TRUE)

model.set=generate.model.set(use.dat=use.dat,
                          test.fit=Model1,
                          pred.vars.cont=cont.preds,
                          pred.vars.fact=cat.preds)
out.list=fit.model.set(model.set, parallel = FALSE, 
                       report.unique.r2 = TRUE)
```

We can examine the output with:

```{r examine}
names(out.list)
out.list$failed.models
length(out.list$success.models)
mod.table=out.list$mod.data.out

```

And examine the full model table.

```{r modtab}
mod.table=mod.table[order(mod.table$AICc),]
tab <- mod.table |>
  as_tibble() |> 
  select(modname, AICc, r2.vals, edf, delta.AICc, wi.AICc)

knitr::kable(
  tab,
  digits = 3,
  caption = "Model comparison summary based on AICc"
)

```

It is always good to check the predictor correlation matrix:

```{r predcors}
model.set$predictor.correlations
```

We can run the same thing using the non.linear correlation matrix, which can be useful if you think there might be strong non-linear dependencies between predictors.

```{r nonlincors}
model.set=generate.model.set(use.dat=use.dat,
                          test.fit=Model1,
                          pred.vars.cont=cont.preds,
                          pred.vars.fact=cat.preds,
                          non.linear.correlations=TRUE)
model.set$predictor.correlations
```

```{r nonlinfit}
out.list=fit.model.set(model.set)
mod.table=out.list$mod.data.out
mod.table=mod.table[order(mod.table$AICc),]
tab <- mod.table |>
  as_tibble() |> 
  select(modname, AICc, r2.vals, edf, delta.AICc, wi.AICc)

knitr::kable(
  tab,
  digits = 3,
  caption = "Model comparison summary based on AICc"
)
```

## Example running across a range of response variables

Set up a vector of response variables

```{r respvec}
resp.vars=c("Acropora.spp.","Turbinaria.spp.","Pocillopora.spp.","Porites.spp.")
```

Get rid of NA's and unused columns

```{r datuse}
use.dat=na.omit(dat[,c(null.vars,cat.preds,cont.preds,resp.vars,"totalpoints")])
```

Now set up some objects to store the results and run the analysis looping through the response variables.

```{r run, include=FALSE}
out.all=list()
var.imp=list()
fss.all=list()
top.all=list()

for(i in 1:length(resp.vars)){
 use.dat$response=use.dat[,resp.vars[i]]
 Model1=uGamm(cbind(use.dat$response,use.dat$totalpoints-use.dat$response)~s(Depth,k=4,bs='cr'),
              family=binomial(), random=~(1|Site),
             data=use.dat,
             lme4=TRUE)

 model.set=generate.model.set(use.dat=use.dat,
                          test.fit=Model1,
                          pred.vars.cont=cont.preds,
                          pred.vars.fact=cat.preds)
 out.list=fit.model.set(model.set)
 fss.all=c(fss.all,list(out.list))
 mod.table=out.list$mod.data.out
 mod.table=mod.table[order(mod.table$AICc),]
 out.i=mod.table
 out.all=c(out.all,list(out.i))
 var.imp=c(var.imp,list(out.list$variable.importance$aic$variable.weights.raw))
 all.less.2AICc=mod.table[which(mod.table$delta.AICc<2),]
 top.all=c(top.all,list(all.less.2AICc))

 # plot the all best models
 par(oma=c(1,1,4,1))
 for(r in 1:nrow(all.less.2AICc)){
 best.model.name=as.character(all.less.2AICc$modname[r])
 best.model=out.list$success.models[[best.model.name]]
 if(best.model.name!="null"){
  plot(best.model$gam,all.terms=T,pages=1,residuals=T,pch=16)
  mtext(side=3,text=resp.vars[i],outer=T)}
 }
}

names(out.all)=resp.vars
names(var.imp)=resp.vars
names(top.all)=resp.vars
names(fss.all)=resp.vars

all.mod.fits=do.call("rbind",out.all)
all.var.imp=do.call("rbind",var.imp)
top.mod.fits=do.call("rbind",top.all)

```
Make a heatmap plot of the variable importance scores.

```{r heatmap}
heatmap.2(all.var.imp,notecex=0.4,  dendrogram ="none",
                     col=colorRampPalette(c("yellow","orange","red"))(30),
                     trace="none",key.title = "",keysize=2,
                     notecol="black",key=T,
                     sepcolor = "black", 
                     Rowv=FALSE,Colv=FALSE)
```

And present all the models, and all the top model tables.

```{r all mods}

tab <- all.mod.fits |>
  rownames_to_column("row_id") |>
  separate(
    row_id,
    into = c("taxon", "model_id"),
    sep = "\\.spp\\.\\."
  ) |> 
  arrange(AICc) |> 
  as_tibble() |> 
  select(taxon, modname, AICc, r2.vals, edf, delta.AICc, wi.AICc)  

knitr::kable(
  tab,
  digits = 3,
  caption = "Model comparison summary based on AICc"
)

```















